---
title: "A Five-Layer Simulation Framework for Bayesian Decision Science"
format:
  html:
    toc: true
    code-fold: false
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

# Overview

This document demonstrates a modular simulation framework for decision science. The simulation separates:

1. Generative world model
2. Observation process
3. Analytical models
4. Decision policies
5. Evaluation metrics

The goal is to show how different analytical assumptions affect business decisions.

# Load Packages

```{r}
library(tidyverse)
library(rjags)
library(coda)

source("world_generation.R")
source("analytical_models.R")
source("policy_evaluation.R")
source("plotting.R")
```

# Five-Layer Architecture

```{mermaid}
%%| echo: false

graph TD

world["Generative World"]
observe["Observation Process"]
analysis["Analytical Models"]
decisions["Decision Policies"]
evaluation["Evaluation"]

world --> observe
observe --> analysis
analysis --> decisions
decisions --> evaluation

```

# Layer 1 — Generative World

The hidden world contains:

- latent customer segments,
- heterogeneous price sensitivity,
- lifetime value,
- and inventory constraints.

```{r}
N_customers <- 50

customer_df <- generate_customer_world(
  N_customers = N_customers,
  seed = 2028
)
```

# Simulate Purchase Opportunities

```{r}
N_offers <- 8

offer_df <- simulate_purchase_opportunities(
  customer_df = customer_df,
  N_offers = N_offers
)
```

# Visualize True Demand Curves

```{r fig.width=11, fig.height=7}
sample_customers <- sample(customer_df$customer_id,30)

plot_true_demand_curves(
  offer_df = offer_df,
  sample_customers = sample_customers
)
```

# Layer 2 — Observation Process

The analyst does not observe:

- latent segment,
- true willingness-to-pay,
- or lifetime value.

The analyst only observes:

- price,
- purchase outcome,
- and customer ID.

```{r}
analyst_df <- observe_analyst_data(offer_df)
```

# Layer 3 — Analytical Models

We compare:

1. Naive aggregate model
2. Overfit personalization
3. Hierarchical Bayesian model

# Naive Aggregate Model

```{r}
naive_df <- fit_naive_aggregate_model(analyst_df)
```

```{r fig.width=10, fig.height=6}
plot_naive_demand_curve(naive_df)
```

# Overfit Personalization

```{r}
customer_summary <- summarize_customer_conversion(analyst_df)
```

# Hierarchical Bayesian Model

```{mermaid}
%%| echo: false

flowchart TD

mu_alpha["μᵅ"]
sigma_alpha["σᵅ"]

mu_beta["μβ"]
sigma_beta["σβ"]

alpha["αⱼ"]
beta["βⱼ"]

price["Price"]
purchase["Purchase"]

mu_alpha --> alpha
sigma_alpha --> alpha

mu_beta --> beta
sigma_beta --> beta

alpha --> purchase
beta --> purchase
price --> purchase
```

# JAGS Model

```{r}
model_string <- hierarchical_demand_model_string()
```

```{r}
bayes_fit <- fit_hierarchical_bayesian_model(
  analyst_df = analyst_df,
  N_customers = N_customers,
  model_string = model_string
)

jags_data <- bayes_fit$jags_data
jags_model <- bayes_fit$jags_model
posterior_samples <- bayes_fit$posterior_samples
posterior_matrix <- bayes_fit$posterior_matrix
mu_alpha_post <- bayes_fit$mu_alpha_post
mu_beta_post <- bayes_fit$mu_beta_post
```

# Layer 4 — Decision Policies

The company seeks to maximize profit subject to inventory constraints.

```{r}
unit_cost <- 5
inventory_limit <- 120
price_grid <- seq(1,35,by = 0.25)
```

# True Profit Curve

```{r}
true_profit_df <- compute_true_profit_curve(
  customer_df = customer_df,
  price_grid = price_grid,
  unit_cost = unit_cost,
  inventory_limit = inventory_limit
)
```

# Naive Profit Curve

```{r}
naive_profit_df <- compute_naive_profit_curve(
  naive_df = naive_df,
  N_customers = N_customers,
  unit_cost = unit_cost,
  inventory_limit = inventory_limit
)
```

# Hierarchical Bayesian Profit Curve

```{r}
bayes_profit_df <- compute_bayes_profit_curve(
  price_grid = price_grid,
  N_customers = N_customers,
  mu_alpha_post = mu_alpha_post,
  mu_beta_post = mu_beta_post,
  unit_cost = unit_cost,
  inventory_limit = inventory_limit
)
```

# Compare Profit Curves

```{r fig.width=12, fig.height=8}
comparison_df <- compare_profit_curves(
  true_profit_df = true_profit_df,
  naive_profit_df = naive_profit_df,
  bayes_profit_df = bayes_profit_df
)

plot_profit_curves(comparison_df)
```

# Layer 5 — Evaluation

We evaluate decisions using regret.

$$
\text{Regret}
=
\Pi(p^*_{true})
-
\Pi(p^*_{estimated})
$$

# Compute Regret

```{r}
regret_results <- compute_regret(
  true_profit_df = true_profit_df,
  naive_profit_df = naive_profit_df,
  bayes_profit_df = bayes_profit_df
)

true_optimal_price <- regret_results$true_optimal_price
naive_optimal_price <- regret_results$naive_optimal_price
bayes_optimal_price <- regret_results$bayes_optimal_price
true_optimal_profit <- regret_results$true_optimal_profit
naive_realized_profit <- regret_results$naive_realized_profit
bayes_realized_profit <- regret_results$bayes_realized_profit
regret_df <- regret_results$regret_df

regret_df
```

# Visualize Regret

```{r fig.width=10, fig.height=6}
plot_regret(regret_df)
```

# Interpretation

This framework separates:

- the hidden world,
- the observation process,
- analytical assumptions,
- operational decisions,
- and evaluation metrics.

This allows us to study:

- omitted-variable bias,
- aggregation bias,
- sparse-data instability,
- partial pooling,
- and decision quality.

The key insight is that analytical assumptions directly affect operational decisions.

# Future Extensions

Possible extensions include:

- dynamic pricing,
- Thompson sampling,
- reinforcement learning,
- competitor reactions,
- nonstationary demand,
- customer churn,
- and causal confounding.

The modular architecture allows each component of the simulation to be independently modified and evaluated.
