---
title: "A Five-Layer Simulation Framework for Bayesian Decision Science"
format:
  html:
    toc: true
    code-fold: false
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

# Overview

This document demonstrates a modular simulation framework for decision science. The simulation separates:

1. Generative world model
2. Observation process
3. Analytical models
4. Decision policies
5. Evaluation metrics

The goal is to show how different analytical assumptions affect business decisions.

# Load Packages

```{r}
library(tidyverse)
library(rjags)
library(coda)

source("R/load_scenario.R")
source("generative_models/world_generation.R")
source("generative_models/dynamic_inventory.R")
source("analytical_models/model_fitting.R")
source("pricing_policies/dynamic_pricing_policies.R")
source("policy_evaluation.R")
source("evaluation/oracle_evaluation.R")
source("evaluation/dynamic_evaluation.R")
source("plotting.R")
```

# Load Scenario

```{r}
scenario_file <- Sys.getenv(
  "SCENARIO_FILE",
  unset = "scenarios/baseline_market.yaml"
)

scenario <- load_scenario(scenario_file)

scenario_summary(scenario)
```

# Five-Layer Architecture

```{mermaid}
%%| echo: false

graph TD

world["Generative World"]
time["Time Evolution"]
inventory["Inventory State"]
observe["Observation Process"]
analysis["Analytical Models"]
decisions["Static Decision Policies"]
dynamic_decisions["Dynamic Pricing Policies"]
evaluation["Evaluation"]
oracle["Oracle Evaluation"]

world --> time
time --> inventory
time --> observe
observe --> analysis
analysis --> decisions
analysis --> dynamic_decisions
decisions --> evaluation
dynamic_decisions --> inventory
inventory --> time
evaluation --> oracle

```

# Layer 1 — Generative World

The hidden world contains:

- latent customer segments,
- heterogeneous price sensitivity,
- lifetime value,
- and inventory constraints.

```{r}
unit_cost <- scenario$unit_cost
inventory_limit <- scenario$inventory_limits$total
price_grid <- scenario_price_grid(scenario)
N_customers <- scenario$market_size

customer_df <- generate_customer_world(
  N_customers = N_customers,
  seed = scenario$random_seed,
  segment_proportions = scenario$customer_segment_proportions,
  price_elasticity_parameters = scenario$price_elasticity_parameters,
  lifetime_value_parameters = scenario$lifetime_value
)

true_demand_curve_df <- generate_true_demand_curve(
  customer_df = customer_df,
  price_grid = price_grid
)
```

# Simulate Purchase Opportunities

```{r}
N_offers <- scenario$simulation_horizon

offer_df <- simulate_purchase_opportunities(
  customer_df = customer_df,
  N_offers = N_offers,
  offer_price_range = scenario$offer_price_range,
  inventory_limit = inventory_limit,
  noise_levels = scenario$noise_levels,
  demand_shock_settings = scenario$demand_shock_settings,
  promotional_effects = scenario$promotional_effects
)
```

# Visualize True Demand Curves

```{r fig.width=11, fig.height=7}
sample_customers <- sample(
  customer_df$customer_id,
  min(30, nrow(customer_df))
)

plot_true_demand_curves(
  offer_df = true_demand_curve_df,
  sample_customers = sample_customers
)
```

# Layer 2 — Observation Process

The analyst does not observe:

- latent segment,
- true willingness-to-pay,
- or lifetime value.

The analyst only observes:

- price,
- promotion status,
- purchase outcome,
- and customer ID.

```{r}
analyst_df <- observe_analyst_data(offer_df)
```

# Layer 3 — Analytical Models

We compare:

1. Naive aggregate model
2. Hierarchical Bayesian model
3. Intentionally misspecified model

# Naive Aggregate Model

```{r}
naive_df <- fit_naive_aggregate_model(analyst_df)
```

```{r fig.width=10, fig.height=6}
plot_naive_demand_curve(naive_df)
```

# Overfit Personalization

```{r}
customer_summary <- summarize_customer_conversion(analyst_df)
```

# Hierarchical Bayesian Model

```{mermaid}
%%| echo: false

flowchart TD

mu_alpha["μᵅ"]
sigma_alpha["σᵅ"]

mu_beta["μβ"]
sigma_beta["σβ"]

alpha["αⱼ"]
beta["βⱼ"]

price["Price"]
inventory["Inventory"]
purchase["Purchase"]

mu_alpha --> alpha
sigma_alpha --> alpha

mu_beta --> beta
sigma_beta --> beta

alpha --> purchase
beta --> purchase
price --> purchase
inventory --> price
```

# JAGS Model

```{r}
model_string <- hierarchical_demand_model_string()
```

```{r}
bayes_fit <- fit_hierarchical_bayesian_model(
  analyst_df = analyst_df,
  model_string = model_string
)

jags_data <- bayes_fit$jags_data
jags_model <- bayes_fit$jags_model
posterior_samples <- bayes_fit$posterior_samples
posterior_matrix <- bayes_fit$posterior_matrix
mu_alpha_post <- bayes_fit$mu_alpha_post
mu_beta_post <- bayes_fit$mu_beta_post
```

# Intentionally Misspecified Model

This model intentionally ignores customer heterogeneity and uses a linear price-only demand surface.

```{r}
misspecified_fit <- fit_misspecified_model(analyst_df)
```

# Inferred Parameters

```{r}
inferred_parameters_df <- summarize_inferred_parameters(
  analyst_df = analyst_df,
  bayes_fit = bayes_fit,
  misspecified_fit = misspecified_fit
)

parameter_comparison_df <- compare_inferred_parameters(
  inferred_parameters_df = inferred_parameters_df,
  customer_df = customer_df
)

parameter_comparison_df
```

# Layer 4 — Decision Policies

The company seeks to maximize profit subject to inventory constraints.

```{r}
unit_cost
inventory_limit
```

# True Profit Curve

```{r}
true_profit_df <- compute_true_profit_curve(
  customer_df = customer_df,
  price_grid = price_grid,
  unit_cost = unit_cost,
  inventory_limit = inventory_limit
)
```

# Naive Profit Curve

```{r}
naive_profit_df <- compute_naive_profit_curve(
  naive_df = naive_df,
  N_customers = n_distinct(analyst_df$customer_id),
  unit_cost = unit_cost,
  inventory_limit = inventory_limit
)
```

# Hierarchical Bayesian Profit Curve

```{r}
bayes_profit_df <- compute_bayes_profit_curve(
  price_grid = price_grid,
  N_customers = n_distinct(analyst_df$customer_id),
  mu_alpha_post = mu_alpha_post,
  mu_beta_post = mu_beta_post,
  unit_cost = unit_cost,
  inventory_limit = inventory_limit
)
```

# Misspecified Profit Curve

```{r}
misspecified_profit_df <- compute_misspecified_profit_curve(
  price_grid = price_grid,
  N_customers = n_distinct(analyst_df$customer_id),
  misspecified_fit = misspecified_fit,
  unit_cost = unit_cost,
  inventory_limit = inventory_limit
)
```

```{r}
model_profit_curves <- list(
  "Naive Aggregate" = naive_profit_df,
  "Hierarchical Bayesian" = bayes_profit_df,
  "Misspecified Linear" = misspecified_profit_df
)
```

# Compare Profit Curves

```{r fig.width=12, fig.height=8}
comparison_df <- compare_profit_curves(
  true_profit_df = true_profit_df,
  naive_profit_df = naive_profit_df,
  bayes_profit_df = bayes_profit_df,
  misspecified_profit_df = misspecified_profit_df
)

plot_profit_curves(comparison_df)
```

# Oracle Evaluation

The oracle evaluation layer can inspect the true latent parameters, demand structure, and customer segments. Analytical models do not receive this information; it is used only to score decisions after each model has selected a price.

```{r}
oracle_truth_summary_df <- summarize_oracle_truth(customer_df)

oracle_truth_summary_df
```

```{r}
oracle_metrics_df <- compute_oracle_decision_metrics(
  true_profit_df = true_profit_df,
  model_profit_curves = model_profit_curves
)

oracle_metrics_df
```

```{r}
calibration_metrics_df <- compute_calibration_metrics(
  analyst_df = analyst_df,
  offer_df = offer_df,
  naive_df = naive_df,
  bayes_fit = bayes_fit,
  misspecified_fit = misspecified_fit
)

calibration_metrics_df
```

```{r}
posterior_coverage_df <- compute_posterior_uncertainty_coverage(
  bayes_fit = bayes_fit,
  customer_df = customer_df
)

posterior_coverage_df
```

# Oracle Comparison Plots

```{r fig.width=12, fig.height=8}
revenue_curve_df <- prepare_revenue_curve_comparison(
  true_profit_df = true_profit_df,
  model_profit_curves = model_profit_curves
)

plot_revenue_curve_comparison(revenue_curve_df)
```

```{r fig.width=10, fig.height=6}
plot_optimal_price_comparison(oracle_metrics_df)
```

```{r fig.width=10, fig.height=6}
plot_regret_distributions(oracle_metrics_df)
```

# Dynamic Inventory-Constrained Pricing

The dynamic simulation evolves inventory over time, lets policies update prices from the current inventory state, applies optional replenishment, and scores each policy against an oracle that can inspect the true demand structure.

```{mermaid}
%%| echo: false

flowchart LR

state["Inventory State t"]
policy["Pricing Policy"]
price["Price t"]
customer["Customer Response"]
shock["Demand Shock"]
sales["Purchases / Sales"]
replenish["Optional Replenishment"]
next_state["Inventory State t + 1"]
oracle_dynamic["Dynamic Oracle Benchmark"]

state --> policy
policy --> price
price --> customer
shock --> customer
customer --> sales
sales --> next_state
replenish --> next_state
next_state --> state
state --> oracle_dynamic
price --> oracle_dynamic

```

```{r}
dynamic_policies <- build_dynamic_pricing_policies(
  estimated_profit_df = bayes_profit_df,
  static_price = select_greedy_price(bayes_profit_df)
)

dynamic_results_df <- simulate_dynamic_pricing_policies(
  customer_df = customer_df,
  policies = dynamic_policies,
  horizon = scenario$simulation_horizon,
  initial_inventory = inventory_limit,
  price_grid = price_grid,
  unit_cost = unit_cost,
  demand_shock_settings = scenario$demand_shock_settings,
  replenishment_settings = scenario$replenishment,
  inventory_behavior = scenario$inventory_behavior,
  seed = scenario$random_seed
) %>%
  compute_dynamic_regret(
    customer_df = customer_df,
    price_grid = price_grid,
    unit_cost = unit_cost
  )

dynamic_policy_summary_df <- summarize_dynamic_policy_performance(
  dynamic_results_df
)

dynamic_policy_summary_df
```

```{r fig.width=11, fig.height=6}
plot_inventory_trajectory(dynamic_results_df)
```

```{r fig.width=11, fig.height=6}
plot_dynamic_prices(dynamic_results_df)
```

```{r fig.width=11, fig.height=6}
plot_cumulative_regret(dynamic_results_df)
```

```{r fig.width=10, fig.height=6}
plot_stockout_frequency(dynamic_results_df)
```

# Layer 5 — Evaluation

We evaluate decisions using regret.

$$
\text{Regret}
=
\Pi(p^*_{true})
-
\Pi(p^*_{estimated})
$$

# Compute Regret

```{r}
regret_results <- compute_regret(
  true_profit_df = true_profit_df,
  naive_profit_df = naive_profit_df,
  bayes_profit_df = bayes_profit_df,
  misspecified_profit_df = misspecified_profit_df
)

true_optimal_price <- regret_results$true_optimal_price
naive_optimal_price <- regret_results$naive_optimal_price
bayes_optimal_price <- regret_results$bayes_optimal_price
misspecified_optimal_price <- regret_results$misspecified_optimal_price
true_optimal_profit <- regret_results$true_optimal_profit
naive_realized_profit <- regret_results$naive_realized_profit
bayes_realized_profit <- regret_results$bayes_realized_profit
misspecified_realized_profit <- regret_results$misspecified_realized_profit
regret_df <- regret_results$regret_df

regret_df
```

# Compare Decisions

```{r}
decision_summary_df <- compute_decision_summary(
  true_profit_df = true_profit_df,
  model_profit_curves = model_profit_curves
)

decision_summary_df
```

# Decision Quality Summary

```{r}
best_decision_models_df <- summarize_best_decision_models(
  oracle_metrics_df
) %>%
  mutate(scenario = scenario$name, .before = model)

best_decision_models_df
```

```{r}
best_model <- best_decision_models_df$model[1]
best_profit_regret <- best_decision_models_df$profit_regret[1]
best_decision_regret <- best_decision_models_df$decision_regret[1]

decision_quality_summary <- tibble(
  scenario = scenario$name,
  best_decision_model = best_model,
  profit_regret = best_profit_regret,
  decision_regret = best_decision_regret,
  explanation = case_when(
    best_model == "Hierarchical Bayesian" ~
      "Partial pooling improves price choice when customer-level heterogeneity matters and sparse observations make fully pooled estimates unstable.",
    best_model == "Naive Aggregate" ~
      "The pooled model is competitive when aggregation error is smaller than posterior uncertainty and the optimal price is broad.",
    best_model == "Misspecified Linear" ~
      "The misspecified model can occasionally choose a reasonable price, but its decision quality is fragile because it ignores nonlinear demand and heterogeneity.",
    TRUE ~
      "Decision quality depends on how closely the estimated revenue curve preserves the oracle-optimal price region."
  )
)

decision_quality_summary
```

# Visualize Regret

```{r fig.width=10, fig.height=6}
plot_regret(regret_df)
```

# Interpretation

This framework separates:

- the hidden world,
- the observation process,
- analytical assumptions,
- operational decisions,
- and evaluation metrics.

This allows us to study:

- omitted-variable bias,
- aggregation bias,
- sparse-data instability,
- partial pooling,
- and decision quality.

The key insight is that analytical assumptions directly affect operational decisions.
The oracle layer makes that comparison explicit: the best analytical model is the one that preserves high-profit decisions under the current scenario, even when its parameter estimates are imperfect.

# Future Extensions

Possible extensions include:

- dynamic pricing,
- Thompson sampling,
- reinforcement learning,
- competitor reactions,
- nonstationary demand,
- customer churn,
- and causal confounding.

The modular architecture allows each component of the simulation to be independently modified and evaluated.
